---
output:
  pdf_document: default
  html_document: default
---
# **Weighted gene co-expression network analysis(WGCNA) **


# Loading required libraries.
```{r message=FALSE}
library(DESeq2)
library(WGCNA)
library(magrittr)
library(ggplot2)
library(genefilter)
```

# Read counts data into CSV file.
```{r}
read_count_data <- function(file_path){
  counts_data <- read.csv(file_path, row.names = 1)
  expression_data <- round(counts_data)
  return(expression_data)
}
expression_matrix <- read_count_data("../Network analysis/Data/GSE183947_fpkm.csv")
head(expression_matrix,2)
```

# Read metadata into CSV file.
```{r}
read_metadata <- function(file_path){
  coldata <- read.csv(file_path, row.names = 1)
  return (coldata)
}
meta_data <- read_metadata("../Network analysis/Data/metadata.csv")
head(meta_data)
```

## Convert condition column in metadata to factor.
```{r}
meta_data$condition <- as.factor(meta_data$condition)
meta_data$description <- as.factor(meta_data$description)
```

## Make sure the row names in metadata matches to the column names in expression matrix.
```{r}
all(rownames(meta_data) %in% colnames(expression_matrix))
```

## Match the row names in metadata to the column names in expression matrix.
```{r}
rownames(meta_data) = colnames(expression_matrix)
```

## Create a new column named accession_code and store the colnames of expression matrix in it 

```{r}
meta_data$accession_code <- colnames(expression_matrix)
```


# pre-filtering to keep only genes with 50 or more reads in total across the samples.

```{r}
pre_filter <- function(){
  # Only keep rows that have total counts above the cutoff
  keep <- expression_matrix %>% rowSums(.) >= 50
  filtered_counts <- expression_matrix[keep,]
  return (filtered_counts)
}
filtered_expression_counts <- pre_filter()
head(filtered_expression_counts,2)
```

# Construct a DESeqDataSet.
```{r}
deseqdataset <- function(){
  deseqdataset <- DESeqDataSetFromMatrix(countData = filtered_expression_counts, 
                                         colData = meta_data, 
                                         design = ~ condition)
  
  return(deseqdataset)
}
deseqdataset_object <- deseqdataset()
deseqdataset_object
```

# Differential expression analysis
```{r}
diff_expr_analysis <- function(){
  deseq_analysis <- DESeq(deseqdataset_object)
  # Apply Variance Stabilizing Transformation (VST)
  vsd <- vst(deseq_analysis, blind = FALSE)
  return (vsd)
}
dds_norm <- diff_expr_analysis()
dds_norm
```

# Extract the VST-transformed data, Filter low variance genes then transpose to have genes as columns
```{r}
norm_counts <- function(){
  norm_transposed_vst <- assay(dds_norm) %>% 
              varFilter(var.cutoff = 0.5) %>% 
              t() 
  return(norm_transposed_vst)
}
normalized_counts <- norm_counts()
```

# Determine power soft-threshold
```{r}
pick_s_th <- function(){
# the pickSoftThreshold() function help identify good choices for power parameter
  sft <- pickSoftThreshold(normalized_counts,
    dataIsExpr = TRUE,
    corFnc = cor,
    networkType = "signed")
  return(sft)
  
}
pick_soft_th <- pick_s_th()
```
## Calculate a measure of the model fit, the signed R^2
```{r}
calculate_model_fit <- function(){
  sft_df <- data.frame(pick_soft_th$fitIndices) %>%
    dplyr::mutate(model_fit = -sign(slope) * SFT.R.sq)
  return(sft_df)
}
model_fit_df <- calculate_model_fit()
model_fit_df
```
## Plot the model fitting by the power soft threshold so we can decide on a soft-threshold for power.
```{r}
plot_model_fit <- function(){
jpeg("../Network analysis/outputs/model_fit.jpeg")
p <- ggplot(model_fit_df, aes(x = Power, y = model_fit, label = Power)) +
     geom_point() +
     # We'll put the Power labels slightly above the data points
     geom_text(nudge_y = 0.1) +
     # We will plot what WGCNA recommends as an R^2 cutoff
     geom_hline(yintercept = 0.80, col = "red") +
    # Just in case our values are low, we want to make sure we can still see the 0.80 level
     ylim(c(min(model_fit_df$model_fit), 1.05)) +
     xlab("Soft Threshold (power)") +
     ylab("Scale Free Topology Model Fit, signed R^2") +
     ggtitle("Scale independence") +
     theme_classic()
print(p)
dev.off()
}

plot_model_fit()


```

# Run WGCNA to find gene co-expression modules using 16 for the power argument
```{r}
run_WGCNA <- function(){
bwnet <- blockwiseModules(
  normalized_counts,
  # What size chunks (how many genes) the calculations should be run in
  maxBlockSize = 2000, 
  # topological overlap matrix
  TOMType = "signed", 
  # soft threshold for network construction
  power = 16, 
  # Let's use numbers instead of colors for module labels
  numericLabels = TRUE, 
  randomSeed = 1234
  )
return(bwnet)
}
bwnet <- run_WGCNA()
```

# Write main WGCNA results to CSV file
```{r}
write_WGCNA <- function(out_path){
  write.csv(bwnet$MEs, out_path)
}
write_WGCNA("../Network analysis/outputs/main_WGCNA_results.csv")
```

# Explore WGCNA results
```{r}
# Explore eigengene modules for each sample
mod_eigengenes<- function(){
  module_eigengenes <- bwnet$MEs
  return(module_eigengenes)
}
module_eigengenes <- mod_eigengenes()
head(module_eigengenes,2)
```

# Which modules have biggest differences across two condition groups?
## Run linear model on each module
```{r}
fit_linear_model <- function(){
  # Create the design matrix from the `condition` variable
  des_mat <- model.matrix(~ meta_data$condition)
  # lmFit() needs a transposed version of the matrix
  fit <- limma::lmFit(t(module_eigengenes), design = des_mat)
  # Apply empirical Bayes to smooth standard errors
  fit <- limma::eBayes(fit)
  return(fit)
}

fit <- fit_linear_model()
```

## Apply multiple testing correction and obtain stats in a dataframe
```{r}
dataframe_stats <- function(){
stats_df <- limma::topTable(fit, number = ncol(module_eigengenes)) %>%
  tibble::rownames_to_column("module")
return(stats_df)
}
stats_df <- dataframe_stats()
stats_df
```

# Module 5 seems to be the most differentially expressed across condition groups
# Let’s make plot of module 5
```{r}

modules_dataframe <- function(){
  module_5 <- module_eigengenes %>%
  tibble::rownames_to_column("accession_code") %>%
  dplyr::inner_join(
    meta_data %>%
      dplyr::select(accession_code, condition),
    by = "accession_code"
  )
  return(module_5)
}
modules_df <- modules_dataframe()
head(modules_df,2)

```
# Boxplot of module 5
```{r}
boxplot_mod_5 <- function(){
jpeg("../Network analysis/outputs/boxplot_of_module_5.jpeg")
p <- ggplot(modules_df,aes(x = condition,
                            y = ME5,
                            color = condition)) +
    # a boxplot with outlier points hidden (they will be in the sina plot)
    geom_boxplot(width = 0.2, outlier.shape = NA) +
    # A sina plot to show all of the individual data points
    ggforce::geom_sina(maxwidth = 0.3) +
    theme_classic()

print(p)
dev.off()  
}
boxplot_mod_5()
```

# Boxplot of module 6
```{r}
boxplot_mod_6 <- function(){
jpeg("../Network analysis/outputs/boxplot_of_module_6.jpeg")
p <- ggplot(modules_df,aes(x = condition,
                            y = ME6,
                            color = condition)) +
    # a boxplot with outlier points hidden (they will be in the sina plot)
    geom_boxplot(width = 0.2, outlier.shape = NA) +
    # A sina plot to show all of the individual data points
    ggforce::geom_sina(maxwidth = 0.3) +
    theme_classic()
print(p)
dev.off()  
}
boxplot_mod_6()

```


# What genes are a part of module 5

## Genes corresponding to each module
```{r}
gene_module <- function(module_name){
 gene_module<- tibble::enframe(bwnet$colors, name = "gene",
                                   value = "module") %>%
# Let's add the `ME` part so its more clear what these numbers are and it matches elsewhere
dplyr::mutate(module = paste0(module_name, module))
 return(gene_module)
}
gene_module_key <- gene_module("ME")
```

## Genes that part of module 5
```{r}
gene_module_key %>% dplyr::filter(module == "ME5")
```

# Extract ME5 eigengene module values 
```{r}
mod_eigengene <- function(module_name){
module_eigengene <- module_eigengenes %>%
    dplyr::select(all_of(module_name)) %>%
    tibble::rownames_to_column("accession_code")

return(module_eigengene)
}
module_eigengene <- mod_eigengene("ME5")
head(module_eigengene,10)
```
# Create dataframe that contain condition and ME5 columns 
```{r}
col_annotation_df <- function(){
  # Set up column annotation from metadata
  col_annot_df <- meta_data %>%
  # Only select the condition and sample ID columns
  dplyr::select(accession_code, condition) %>%
  # Add on the eigengene expression by joining with sample IDs
  dplyr::inner_join(module_eigengene, by = "accession_code") %>%
  # Arrange by condition
  dplyr::arrange(condition) %>%
  # Store sample
  tibble::column_to_rownames("accession_code")
 
  return(col_annot_df) 
}

col_annot_df <- col_annotation_df()
head(col_annot_df,10)
```
# Create the ComplexHeatmap column annotation function
```{r}
ComplexHeatmap_col_annotation <- function(module_name){
  # Create the ComplexHeatmap column annotation object
  col_annot <- ComplexHeatmap::HeatmapAnnotation(
  # Supply condition labels
  condition = col_annot_df$condition,
  # Add annotation barplot
  module_eigengene = ComplexHeatmap::anno_barplot(dplyr::select(col_annot_df, module_name)),
  # Pick colors for each experimental group in condition
  col = list(condition = c("tumor" = "#f1a340", "normal" = "#998ec3"))
  )
  
  return(col_annot)
}

col_annot <- ComplexHeatmap_col_annotation("ME5")
```

# Get a vector of the gene IDs that correspond to this module
```{r}
get_module_genes <- function(module_name){
  module_genes <- gene_module_key %>%
      dplyr::filter(module == module_name) %>%
      dplyr::pull(gene)
  return(module_genes)
}

module_genes <- get_module_genes("ME5")
```

# Set up the gene expression data frame
```{r}
module_matrix <- function(){
mod_mat <- normalized_counts %>%
    t() %>%
    as.data.frame() %>%
    # Only keep genes from ths module
    dplyr::filter(rownames(.) %in% module_genes) %>%
    # Order the samples to match col_annot_df
    dplyr::select(rownames(col_annot_df)) %>%
    # Data needs to be a matrix
    as.matrix()
return(mod_mat)
}

module_mat <- module_matrix()
```

# Normalize the gene expression values
```{r}
norm_module_matrix <- function(){
  mod_mat <- module_mat %>%
    # Scale can work on matrices, but it does it by column so we will need to
    # transpose first
    t() %>%
    scale() %>%
    # And now we need to transpose back
    t()
return(mod_mat)  
}

mod_mat <- norm_module_matrix()
```

# Create a color function based on standardized scale
```{r}
color <- function(){
  color_func <- circlize::colorRamp2(c(-4, 0, 4), 
                                     c("#67a9cf", "#f7f7f7", "#ef8a62"))
  return(color_func)
}

color_func <- color()
```

# Plot the Heatmap
```{r}
plot_heatmap <- function(module_name){
  set.seed(432)
  jpeg("../Network analysis/outputs/Heatmap_of_largest_DE_Module.jpeg")
  heatmap <- ComplexHeatmap::Heatmap(mod_mat,
    name = module_name,
    # Supply color function
    col = color_func,
    # Supply column annotation
    bottom_annotation = col_annot,
    # We don't want to cluster samples
    cluster_columns = FALSE,
    # We don't need to show sample or gene labels
    show_row_names = FALSE,
    show_column_names = FALSE
  )
  print(heatmap)
  dev.off()
}  
plot_heatmap("ME5")
```










